---
title: 'Movielense Data: Recommender System'
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    theme: readable
    toc: yes
    toc_float: yes
date: "CS598, Fall 2020"
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, message=FALSE}
library(recommenderlab)
library(Matrix)
library(data.table)
```

## Read in Data

```{r}
myurl = "https://liangfgithub.github.io/MovieData/"

#ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
#                   sep = ':',
#                   colClasses = c('integer', 'NULL'), 
#                   header = FALSE)
#colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
#ratings$Timestamp = NULL

movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Year = as.numeric(unlist(lapply(movies$Title, 
                                       function(x) substr(x, nchar(x)-4, nchar(x)-1))))

total <- merge(movies,ratings,by="MovieID")
genres = 	c("Action",
	          "Adventure",
	          "Animation",
	          "Children's",
	          "Comedy",
	          "Crime",
            "Documentary",
	          "Drama",
	          "Fantasy",
	          "Film-Noir",
	          "Horror",
	          "Musical",
	          "Mystery",
	          "Romance",
	          "Sci-Fi",
	          "Thriller",
	          "War",
	          "Western")


```


## Training Test Split

Create

* **train** data that contain about 80\% rows of `ratings.dat` 

* **test** data that contain about 20\% of rows of `ratings.dat`


```{r}
set.seed(100)
train.id = sample(nrow(ratings), floor(nrow(ratings)) * 0.8)
train = ratings[train.id, ]
head(train)

test = ratings[-train.id, ]
head(test)
```

## Recommender System

* First create a utility matrix stored as a sparse matrix. 

```{r}
i = paste0('u', train$UserID)
j = paste0('m', train$MovieID)
x = train$Rating
tmp = data.frame(i, j, x, stringsAsFactors = T)
Rmat = sparseMatrix(as.integer(tmp$i), as.integer(tmp$j), x = tmp$x)
rownames(Rmat) = levels(tmp$i)
colnames(Rmat) = levels(tmp$j)
Rmat = new('realRatingMatrix', data = Rmat)
```

`Rmat` is a 6040-by-3681 sparse matrix. Its rows correspond to the unique 6040 users in the training data, and columns correspond to the unique 3681 movies  in the training data (although the MovieIDs range from 1 to 3952). 

```{r}
range(train$MovieID)
length(unique(train$MovieID))
range(train$UserID)
length(unique(train$UserID))
dim(Rmat)
```


* Train a recommender system using R package [[recommenderlab](https://cran.r-project.org/web/packages/recommenderlab/vignettes/recommenderlab.pdf)]

```{r}
rec_UBCF = Recommender(Rmat, method = 'UBCF',
                  parameter = list(normalize = 'Z-score', 
                                   method = 'Cosine', 
                                   nn = 25))
#Summary of model parameters
rec_UBCF@model
```

* Make prediction on the test data

First let's predict all un-rated movies for the first three users in the training data. No information for users not in the training data; the coldstart issue. 

```{r}
# predict ratings for the first three users
recom = predict(rec_UBCF, 
                Rmat[1:3], type = 'ratings')
```

We display the original and predicted ratings of the first 10 movies for the three users. Note that the final prediction could still have missing values. 

```{r}
as(recom, 'matrix')[, 1:10]
as(Rmat, 'matrix')[1:3, 1:10]
```

You can use the code blow to compute RMSE for the test data. Note that the prediction   takes a long time. You could also use the built-in evaluation functions in `recommenderlab`; check Chap 3 of [[recommenderlab](https://cran.r-project.org/web/packages/recommenderlab/vignettes/recommenderlab.pdf)].

```{r, eval=FALSE}
# This may take a long time
recom = predict(rec_UBCF, Rmat, type = 'ratings')  
rec_list = as(recom, 'list')  # each element are ratings of that user

test.pred = test
test.pred$rating = NA

# For all lines in test file, one by one
for (u in 1:nrow(test)){

    # Read userid and movieid from columns 2 and 3 of test data
    userid = paste0('u', test$UserID[u])
    movieid = paste0('m', test$MovieID[u])
    
    rating = rec_list[[userid]][movieid]
    # handle missing values; 2.5 might not be the ideal choice
    test.pred$rating[u] = ifelse(is.na(rating), 2.5, rating)
}

# Calculate RMSE
sqrt(mean((test$Rating - test.pred$rating)^2)) 
```

All algorithms available in `recommenderlab`. Check  [[recommenderlab](https://cran.r-project.org/web/packages/recommenderlab/vignettes/recommenderlab.pdf)] for technical details. 

```{r}
recommenderRegistry$get_entry_names()
```

