---
title: "PSL (F20) Coding Assignemnt 3"
output: 
   html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Frederick (Eric) Ellwanger - fre2
#### 10/13/2020  

### Introduction
Loess short for Local Regression is a non-parametric approach that fits multiple regressions in local neighborhood. This can be particularly resourceful, if you know that your X variables are bound within a range.    
Loess regression can be applied using the loess() function on a numerical vector to smoothen it and to predict the Y locally (i.e, within the trained values of Xs). The size of the neighborhood can be controlled using the span argument, which ranges between 0 to 1. It controls the degree of smoothing. So, the greater the value of span, the more smooth is the fitted curve.

The predictor variable can just be indices from 1 to the number of observations in the absence of explanatory variables. If other explanatory variables are available, they can be used as well (maximum of 4).


# Prepare Function  
```{r}
lo.lev <- function(x1, sp){
  # x1: feature vector of length n
  # sp: a value for "span"
  n = length(x1);
  lev = rep(0, n)
  
  ##############################################
  # YOUR CODE: Compute the diagonal entries of the 
  #            smoother matrix S and 
  #            store it in a vector "lev"
  # Tip: check how we compute the smoother matrix
  #      for smoothing spline models
  ##############################################
  A = matrix(0, n, n);
  for(i in 1:n){
       y = rep(0, n); y[i]=1;
       yi = loess(y ~ x1, span=sp, control = loess.control(surface = "direct"))$fitted;
       A[,i]= yi;
  }
 
  return(diag((A+t(A))/2))
}

onestep_CV <- function(x1, y1, sp){
  
  ##############################################
  #  YOUR CODE: 
  #  1) Fit a loess model y1 ~ x1 with span = sp, and extract 
  #     the corresponding residual vector
  #  2) Call lo.lev to obtain the diagonal entries of S
  #  3) Compute LOO-CV and GCV using formula from lecture notes
  #    [lec_W5_NonlinearRegression.pdf] page 33. 
  ##############################################
  lofit = loess(y1 ~ x1, span = sp, control = loess.control(surface = "direct"))
  resid = lofit$residuals
  lev = lo.lev(x1, sp)

  cv = sum(((resid)/(1-lev))^2)/lofit$n
  gcv = sum(((resid)/(1-sum(lev)/lofit$n))^2)/lofit$n
  
  return(list(cv = cv, gcv = gcv))
}

myCV <- function(x1, y1, span){
  # x1: feature vector of length n
  # y1: response vector of length n
  # span: a sequence of values for "span"
  
  m = length(span)
  cv = rep(0, m)
  gcv = rep(0, m)
  
  for(i in 1:m){
    tmp = onestep_CV(x1, y1, span[i])
    cv[i] = tmp$cv
    gcv[i] = tmp$gcv
  }
  return(list(cv = cv, gcv = gcv))
}
```  


# Test Function

Test function with data loaded from disk  
```{r}
mydata = read.csv(file = "Coding3_Data.csv")
dim(mydata)
```  


Take a look at the plot of the data  
```{r}
plot(mydata$x, mydata$y, xlab = '', ylab = '')
```  


Create a grid of values for span:  15 values equally spaced between 0.20 and 0.90  
```{r}
span1 = seq(from=0.2, by=0.05, length=15)
cv.out = myCV(mydata$x, mydata$y, span1)
myout = data.frame(CV = cv.out$cv, GCV = cv.out$gcv, span = span1)
```  

# Results  
Compute LOO-CV and GCV using function above  
```{r}
myout
```  

Find value for span that creates lowest RSS for both GCV  
```{r}
(msgcv = myout$span[myout$GCV == min(myout$GCV)])
```  


and LOO-CV
```{r}
(mscv = myout$span[myout$CV == min(myout$CV)])
```  


# Plot the fitted curve  
Show the data (red circles), the true curve (grey line), and the fitted curve (blue dashed line) using optimal span.  
```{r}
#Plot data with min GCV value
spangcv.min = msgcv
plot(mydata$x, mydata$y, xlab='', ylab='', col='red')

#Plot true curve 
fx = 1:50/50
fy = sin(12*(fx+0.2))/(fx+0.2)
lines(fx, fy, col='grey', lwd=2)

f = loess(y ~ x, data=mydata, span=spangcv.min)
lines(fx, predict(f, data.frame(x=fx), surface='direct'), lty=2, lwd=2, col='blue')
```   









